Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2025-04-29 manuscript.Rmd
remove si

journal: "Agroforestry systems"
date: ""
output:
  officedown::rdocx_document: 
    base_format: "bookdown::word_document2"
    reference_docx: misc/word-styles-ref-02.docx
bibliography: misc/references.bib
nocite: |
  @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{r setup, include=FALSE}
# Run with R version 4.4.1
# package management is done with renv
packages_needed <- c(
  "ggplot2", "ggrepel",  "ggspatial",  "ggpubr",  "ggfortify",  "maps", 
  "rdryad", "readxl", "data.table", "knitr", "BIOMASS", "terra", "car", 
  "flextable"
)

lapply(packages_needed, require, character.only = TRUE)

@@

```{r tab.id = "abbrev", results = "show", tab.cap = "List of abbreviations used in the paper."}
data.frame(
  Abbreviation = c("AFS", "AGC", "BGC", "CBH", "CF", "FG", "HG", "SOC", "SP", "YA"),
  Definition = c(
    "Agroforestry system", "Aboveground carbon", "Belowground carbon",
    "Circumference at breast height", "Crop and fallow", "Forest garden",
    "Home garden", "Soil organic carbon", "Silvopastoral",
    "Young agroforest"
  )
) |>
  flextable(cwidth = c(1, 2))
```

@@

# Supplemental Information

## Supplemental methods

### Calculation of 10-20 cm average carbon content

During field data collection, because the auger length (15 cm) didn't
match the target soil sampling depths (10 cm each), we could only
accurately sample and analyse the 0-10 cm and 20-30 cm soil depths for
SOC content. The SOC content of the intermediate 10-20 cm soil depth had
to be estimated based on these first two measurements.

To reduce the bias in this estimation, we used the standard assumption
that SOC content $c(x)$ decreases as an exponential function of soil
depth $x$:

$$c(x) = C_0 \cdot exp(-\lambda\cdot x) $$ where $C_0$ is the carbon
content at the surface (depth $x= 0$) and $\lambda$ is the rate at which
the carbon content decreases with soil depth. Under this assumption, we
can calculate the mean SOC content between depths $x$ and $x+\Delta$ as:

$$\bar c(x, x+\Delta) = \frac{1}{\Delta}\int_{x}^{x+\Delta}  C_0 \cdot exp(-\lambda\cdot x) dx$$

$$ \Rightarrow \bar c(x, x+\Delta) = \frac{C_0}{-\lambda \cdot \Delta} \left[ exp(-\lambda\cdot (x+\Delta)) - exp(-\lambda\cdot x) \right]$$

$$ \Rightarrow \bar c(x, x+\Delta) = \frac{C_0}{\lambda \cdot \Delta} exp(-\lambda\cdot x)\left[1- exp(-\lambda\cdot\Delta) \right]$$
The average SOC contents between 0-10 cm, 10-20 cm and 20-30 cm can
respectively be written as:

$$C_{0,10} = \frac{C_0}{10\lambda}\left(1- exp(-10\lambda) \right)$$

$$C_{10,20} = \frac{C_0}{10\lambda} exp(-10\lambda)\left(1- exp(-10\lambda) \right)$$

$$C_{20,30} = \frac{C_0}{10\lambda} exp(-20\lambda)\left(1- exp(-10\lambda) \right)$$
We can thus write:

$$\frac{C_{20,30}}{C_{0,10}} = \frac{\frac{C_0}{10\lambda} exp(-20\lambda)\left(1- exp(-10\lambda) \right)}{\frac{C_0}{10\lambda}\left(1- exp(-10\lambda) \right)} = exp(-20\lambda)$$
$$ \Rightarrow \lambda = \frac{log(\frac{C_{0,10}}{C_{20,30}})}{20}$$ We
can also write:

$$ \frac{C_{10,20}}{C_{0,10}} = exp(-10\lambda) $$
$$\Rightarrow C_{10,20} = C_{0,10} \cdot exp(-10 \cdot \frac{log(\frac{C_{0,10}}{C_{20,30}})}{20})$$

$$ \Rightarrow C_{10,20} = C_{0,10} \cdot \left( \frac{C_{20,30}}{C_{0,10}}\right)^{\frac{1}{2}}$$
### Choice of climate and soil covariates

As described in the "Carbon stocks model" section of the Materials and
Methods, we hypothesized that C stocks (both AGC and SOC) are affected
by climate and soil characteristics, together with land use (in this
case, AFS type) and other spatial differences (in our case, a site
effect).

```{r pca-clim-soil}
pca_clim <- data_plot[, c("temp", "prec", "varp")] |>
  scale() |>
  princomp()

data_soil_melt <-
  data_soil |>
  melt(
    id.vars = c("N_PLOT", "SAMPLE_DEPTH"),
    measure.vars = c(
      "pH\n", "CEC\n", "CLAY", "FINE SILT",
      "COARSE SILT", "FINE SAND", "COARSE SAND"
    )
  )

# make values numeric; when detectability limit is reached, used limit value
data_soil_melt[grepl("LD", value), value := gsub(",", ".", tstrsplit(value, "LD=")[[2]])]

data_soil_melt <-
  data_soil_melt[, .(value = mean(as.numeric(value), na.rm = TRUE)), .(N_PLOT, variable)] |>
  dcast(N_PLOT ~ variable)
data_soil_melt[, silt := `COARSE SILT` + `FINE SILT`]
data_soil_melt[, sand := `FINE SAND` + `COARSE SAND`]

pca_soil <- data_soil_melt[, c("pH\n", "CLAY", "silt", "sand")] |>
  scale() |>
  princomp()
```

We selected the following set of variables to represent the effect of
climate: mean annual temperature (BIO1), mean annual precipitation
(BIO12), annual precipitation variability (BIO15) over the period
1981-2010. These bioclimatic variables were extracted from CHELSA V2.1
[@Karger2017]. These variables were all highly correlated in our dataset
(Figure S\@ref(fig:choice-clim-vars)a). We performed a principal
component analysis after scaling and centering these three variables:
the first component explained
`r round(pca_clim$sdev[1]^2/sum(pca_clim$sdev^2)*100)`% of the variance
in our dataset (Figure S\@ref(fig:choice-clim-vars)b). We therefore
decided to keep only one variable to represent this climatic gradient:
mean annual temperature, which was most correlated with the first
component of the PCA (Figure S\@ref(fig:choice-clim-vars)b), and has
been shown to influence C stocks in other contexts [@Wiesmeier2019;
@Muller-Landau2021].

We selected the following set of variables to represent the effect of
soil properties on C stocks: clay content (\<2 $\mu$m), silt content
(2-50 $\mu$m), sand content (50-2000 $\mu$m), and pH. These variables
were measured on our soil samples analysed in the Cirad laboratories
(see section "Field inventories"). After scaling and centering these
variables, we performed a principal component analysis: the first
component explained
`r round(pca_soil$sdev[1]^2/sum(pca_soil$sdev^2)*100)`% of the variance
in our dataset (Figure S\@ref(fig:choice-soil-vars)b). We therefore
decided to retain only one variable to represent this edaphic gradient:
clay content, which was the most highly correlated with the first
component of the PCA (Figure S\@ref(fig:choice-soil-vars)b) and has been
shown to be a good predictor of SOC stocks in other contexts
[@Wiesmeier2019].

## Supplemental tables

```{r site-soil-char}
data_soil[, Corg := `Corg\n`]
data_soil[, muCorg := mean(Corg, na.rm = TRUE), .(N_PLOT)]
data_soil[is.na(Corg), Corg := muCorg]

data_soil[, c("h0", "h1") := lapply(tstrsplit(SAMPLE_DEPTH, "-"), as.numeric)]

data_soil$silt <- as.numeric(data_soil$`FINE SILT`) + as.numeric(data_soil$`COARSE SILT`)
data_soil$sand <- as.numeric(data_soil$`FINE SAND`) + as.numeric(data_soil$`COARSE SAND`)

# add site info to data_soil
data_soil_plot <- merge(data_soil, data_plot[, c("N_PLOT", "site")])

# choose variables to show
vars <- c("Corg\n", "N", "Corg_N", "TC", "CLAY", "silt", "sand", "pH\n", "Ca", "Mg\n", "K", "Na", "CEC\n")

data_soil_plot <- melt(data_soil_plot,
  id.vars = c("N_PLOT", "SAMPLE_DEPTH", "site"),
  measure.vars = vars
)

data_soil_plot[, variable := gsub("\n", "", variable)]

# when value below detection limit: use half of detection value
data_soil_plot[grepl("LD", value), value := as.numeric(strsplit(value, "=")[[1]][2]) / 2]
data_soil_plot[, value := as.numeric(value)]

# remove NA values
data_soil_plot <- subset(data_soil_plot, !is.na(value))

# simulation 1000 values per plot, soil depth and variable with a 12% coefficient
# of variation (CV value from lab)
data_soil_plot_simu <-
  data_soil_plot[
    , .(
      value = rnorm(1000, value, sd = 0.12 * value), n = 1:1000
    ),
    .(site, N_PLOT, SAMPLE_DEPTH, variable)
  ]

# bootstrap plot values to get mean value at the site level (for each simulation)
data_soil_site_simu <-
  data_soil_plot_simu[
    , .(
      value = mean(sample(value, length(value), replace = TRUE))
    ),
    .(site, variable, n)
  ]

# convert sand, silt, clay and C content from % into g/kg
data_soil_site_simu[variable %in% c("Corg", "CLAY", "sand", "silt", "TC"), value := value * 10]

# calculate mean and 95% confidence interval at site level
data_soil_site <- data_soil_site_simu[
  , .(
    mean = signif(mean(value), 3),
    lwr = signif(quantile(value, 0.025), 3),
    upr = signif(quantile(value, 0.975), 3)
  ),
  .(site, variable)
]
data_soil_site[, value := paste0(round(mean, 1), " (", round(lwr, 1), "-", round(upr, 1), ")")]
data_soil_site[variable %in% c("K", "Mg", "Na"), value := paste0(round(mean, 2), " (", round(lwr, 2), "-", round(upr, 2), ")")]

# change table shape to have one column per variable
data_soil_site <- dcast(data_soil_site, site ~ variable, value.var = "value")

## add unit row
data_soil_site <- data_soil_site[, c("site", "CLAY", "silt", "sand", "Corg", "N", "Corg_N", "TC", "pH", "CEC", "Ca", "K", "Mg", "Na")]
colnames(data_soil_site) <- c(" ", "Clay", "Silt", "Sand", "Organic C content", "Total N content", "C:N", "Carbonate content", "pH", "CEC", "Ca", "K", "Mg", "Na")
```

```{r tab.id = "site-soil", tab.lp="supp-tab", results = "show", tab.cap.pre="Table S", tab.cap = "Soil characteristics of the sites (average over 0-30 cm), presented as mean values with 95% confidence intervals (in parentheses). Units are: in g/kg for clay, silt, sand, carbonate, organic C and total N contents; in cmolc/kg for CEC, Ca, K, Mg and Na; and no units for C:N and pH. The confidence intervals were calculated by propagating a 12% error on the lab-provided values and obtaining site-level estimates through bootstrapping plot values (1000 simulations). Organic carbon content (difference of the total carbon – dry combustion method – and the carbonates contents - calcimeter) ; total nitrogen content (dry combustion method) ; pH in water (soil:deionized water ratio of 1:2.5) ; CEC - cation exchangeable capacity - and exchangeable cations (Ca, K, Mg and Na) at soil pH [Cobaltihexamine method; Orsini and Remy, 1976]."}
flextable(data_soil_site, cwidth = 0.5)
```

```{r tab.id = "afs-typology", tab.cap.pre="Table S", tab.cap = "Description of 5 agroforestry systems in Central Timor-Leste.", tab.autonum.start_at=1, tab.lp="supp-tab", results = "show"}
read_excel("tables/afs-typology.xlsx") |>
  flextable(
    col_keys = c(
      "Typology adapted from Nair (1993)",
      "Usual location",
      "Typical plot size (ha)",
      "Time since installation",
      "Previous land use"
    ),
    cwidth = c(1.5, 1.5, 1, 1, 1.5)
  )
```

```{r tetum-scientific-correspondance}
sheets1 <- grep("INV", excel_sheets("data/DATA_BIOMASS_TL_LARGE.xlsx"), value = TRUE)
sheets2 <- grep("INV", excel_sheets("data/DATA_BIOMASS_TL_SHORT.xlsx"), value = TRUE)

vern1 <- lapply(sheets1, \(x){
  df <- read_excel("data/DATA_BIOMASS_TL_LARGE.xlsx", sheet = x)
  df <- df[, c("TETUM_NAME", "SCIENTIFIC_NAME")] |>
    unique() |>
    subset(!SCIENTIFIC_NAME %in% c("Unknown", "NR", "Unkown"))
})
vern2 <- lapply(sheets2, \(x){
  df <- read_excel("data/DATA_BIOMASS_TL_LARGE.xlsx", sheet = x)
  df <- df[, c("TETUM_NAME", "SCIENTIFIC_NAME")] |>
    unique() |>
    subset(!SCIENTIFIC_NAME %in% c("Unknown", "NR", "Unkown"))
})

vern <- rbind(rbindlist(vern1), rbindlist(vern2)) |> unique()
vern <- subset(vern, !is.na(TETUM_NAME) & !is.na(SCIENTIFIC_NAME))

## correct typos in vernacular names
vern[, TETUM_NAME := tolower(gsub(" oan| kain| restu| hun restu", "", TETUM_NAME))]
vern[grep("kami", TETUM_NAME), TETUM_NAME := "ai-kamii"]
vern[, TETUM_NAME := gsub("ai-haek", "ai-hanek", TETUM_NAME)]
vern[, TETUM_NAME := gsub("ai-bobur", "ai-bubur", TETUM_NAME)]
vern[, TETUM_NAME := gsub("duut", "du'ut", TETUM_NAME)]
vern[, TETUM_NAME := gsub("aifarina", "ai-farina", TETUM_NAME)]
vern[, TETUM_NAME := gsub("aitarak", "ai-tarak", TETUM_NAME)]
vern[, TETUM_NAME := gsub("ainanas", "ai-nanas", TETUM_NAME)]
vern[, TETUM_NAME := gsub("fore tali", "fore-tali", TETUM_NAME)]
vern[, TETUM_NAME := gsub("kromolaina", "kromolaena", TETUM_NAME)]

## correct scientific names
vern[, c("genus", "species") := tstrsplit(SCIENTIFIC_NAME, " ")[1:2]]
vern[species %in% c("sp", "sp."), species := NA]
vern[species == "purpuroides", species := NA]
corr <- correctTaxo(genus = vern$genus, species = vern$species)
vern <- cbind(vern, corr)

vern[, latin := paste(genusCorrected, speciesCorrected)]
vern[, latin := gsub(" NA", " sp.", latin)]
vern[, tetum := gsub("  ", " ", tolower(TETUM_NAME))]
vern <- vern[, c("tetum", "latin", "genusCorrected", "speciesCorrected")]
vern <- unique(vern)

# check that all names are unique
# if one tetum name has two levels of id (same genus), keep better identification
dupl_tetum <- vern[duplicated(tetum), unique(tetum)]
vern <- vern[!(tetum %in% dupl_tetum & is.na(speciesCorrected))]
setorder(vern, latin)

vern <- vern[, .(tetum = paste(tetum, collapse = ", ")), .(latin)]
```

```{r tab.id = "tetum-names", tab.lp="supp-tab", results = "show", tab.cap.pre="Table S", tab.cap = "Correspondence between Tetum names and scientific names used in this study."}
flextable(vern) |>
  set_header_labels(values = c("Scientific name", "Tetum name")) |>
  autofit() |>
  italic(j = 1)
```

## Supplemental figures

```{r fig.id="design", fig.cap = "Plot sampling design. (a) Typical plot < 1 ha, in all AFS except SPs. (b) Typical 1-ha plot in SPs. Red areas are the five 10x10 m subplots, and blue dots are the three soil sampling locations.", fig.autonum.start_at=1, fig.lp="supp-fig", results = "show", fig.cap.pre="Figure S", fig.width = 5, fig.height = 3}
include_graphics("figures/plot-design.jpg")
```

```{r fig.id="seq-diagram", fig.cap = "Typical sequence of AFS in the sites included in this study. Transitions between two systems are represented by arrows: dashed arrows represent transitions that usually include tree removal, whereas solid arrows represent transitions where trees are planted and/or the AFS is left to regenerate naturally. These transitions may vary according to the farmer's strategy and available resources (Cogné and Lescuyer, 2024).", fig.lp="supp-fig", results = "show", fig.cap.pre="Figure S", fig.width = 8, fig.height = 2.2, out.width="50%"}
# save_png <- function(plot, path){
#   DiagrammeRsvg::export_svg(plot) %>%
#     charToRaw() %>%
#     rsvg::rsvg() %>%
#     png::writePNG(path)
# }
#
# DiagrammeR::grViz("
#       digraph {
#       graph [layout = dot, rankdir = LR]
#
#       node [shape = ellipse]
#       CF[color = '#1B9E77'];
#       SP[color = '#D95F02'];
#       YA[color = '#7570B3'];
#       HG[color = '#E7298A'];
#       FG[color = '#E6AB02'];
#
#       NE[label = 'Forest or\nsavanna', shape = rectangle, fillcolor ='black'];
#
#       # Connect nodes with edges and labels
#       CF -> YA -> HG -> FG;
#       YA -> FG;
#       edge[style = 'dashed']
#       SP -> CF;
#       FG -> CF
#       FG -> YA
#       NE -> SP
#       NE -> CF
#        }
#
#       ") %>%
#   save_png("figures/seq_diagram.png")
include_graphics("figures/seq_diagram.png")
```

<!-- </br> -->

<!-- ```{r fig.id = "dag", fig.cap="Direct acyclic plot of the hypothesized effect of soil characteristics, climate and land use (here AFS) on AGC and SOC stocks.", fig.lp="supp-fig", results = "show", fig.cap.pre="Figure S"} -->

<!-- # save_png <- function(plot, path){ -->

<!-- #   DiagrammeRsvg::export_svg(plot) %>% -->

<!-- #     charToRaw() %>% -->

<!-- #     rsvg::rsvg() %>% -->

<!-- #     png::writePNG(path) -->

<!-- # } -->

<!-- #  -->

<!-- # DiagrammeR::grViz(" -->

<!-- #       digraph { -->

<!-- #       graph [layout = dot, rankdir = LR] -->

<!-- #  -->

<!-- #       node [shape = ellipse] -->

<!-- #       AGC; -->

<!-- #       SOC; -->

<!-- #       node [shape = rectangle] -->

<!-- #       climate; -->

<!-- #       soil; -->

<!-- #       AFS; -->

<!-- #  -->

<!-- #       # Connect nodes with edges and labels -->

<!-- #       AFS -> AGC; -->

<!-- #       climate -> AGC; -->

<!-- #       AFS -> SOC; -->

<!-- #       soil -> AGC; -->

<!-- #       soil -> SOC; -->

<!-- #       climate -> SOC; -->

<!-- #       edge[style = 'dashed'] -->

<!-- #       AGC -> SOC; -->

<!-- #        } -->

<!-- #  -->

<!-- #       ") %>% -->

<!-- #   save_png("figures/dag.png") -->

<!-- include_graphics("figures/dag.png")  -->

```{r fig.id="choice-clim-vars", fig.width = 10, fig.height=4, fig.cap="(a) Pearson coefficient of correlation matrix and (b) principal component analysis biplot, with three climate variables: mean annual temperature (MAT), mean annual precipitation (MAP) and precipitation variability (PVar). Values for the climate variables were extracted from CHELSA V2.1 (Karger et al., 2017) for each of the 30 study plots.", fig.lp="supp-fig", results = "show", fig.cap.pre="Figure S"}
g1 <- data_plot[, c("temp", "prec", "varp")] |>
  setnames(c("MAT", "MAP", "Pvar")) |>
  cor() |>
  melt() |>
  ggplot(aes(x = Var1, y = Var2, label = round(value, 2), fill = value)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradientn(colours = c("brown1", "#FFFFFF", "cornflowerblue")) +
  theme_bw() +
  labs(x = NULL, y = NULL, fill = "Pearson\nCorrelation\nCoefficient")

g2 <- data_plot[, c("temp", "prec", "varp")] |>
  setnames(c("MAT", "MAP", "Pvar")) |>
  scale() |>
  princomp() |>
  autoplot(
    data = data_plot, colour = "site", loadings.colour = "black",
    loadings.label = TRUE, loadings.label.col = "black"
  ) +
  theme_classic()

ggpubr::ggarrange(g1, g2, labels = "auto")
```

```{r fig.id = "choice-soil-vars", fig.width = 10, fig.height=4, fig.cap="(a) Pearson coefficient of correlation matrix and (b) principal component analysis biplot, with four soil variables: clay, silt and sand content, and soil pH. Values for the soil variables were measured from in situ soil samples for each of the 30 study plots.", fig.lp="supp-fig", results = "show", fig.cap.pre="Figure S"}
g1 <- data_soil_melt[, c("CLAY", "silt", "sand", "pH\n")] |>
  setnames(c("clay", "silt", "sand", "pH")) |>
  cor() |>
  melt() |>
  ggplot(aes(x = Var1, y = Var2, label = round(value, 2), fill = value)) +
  geom_tile() +
  geom_text() +
  scale_fill_gradientn(colours = c("brown1", "#FFFFFF", "cornflowerblue")) +
  theme_bw() +
  labs(x = NULL, y = NULL, fill = "Pearson\nCorrelation\nCoefficient")

g2 <- data_soil_melt[, c("CLAY", "silt", "sand", "pH\n")] |>
  setnames(c("clay", "silt", "sand", "pH")) |>
  scale() |>
  princomp() |>
  autoplot(
    data = data_plot, colour = "site", loadings.colour = "black",
    loadings.label = TRUE, loadings.label.col = "black"
  ) +
  theme_classic()
ggpubr::ggarrange(g1, g2, labels = "auto")
```

<!-- ```{r dag-code} -->

<!-- # dag { -->

<!-- # bb="-2.799,-2.922,2.216,3.074" -->

<!-- # afs [exposure,pos="0.265,-1.155"] -->

<!-- # agc [exposure,pos="-0.591,0.840"] -->

<!-- # climate [adjusted,pos="-1.049,-1.155"] -->

<!-- # soc [outcome,pos="0.137,1.233"] -->

<!-- # soil [adjusted,pos="-0.426,-1.182"] -->

<!-- # afs -> agc -->

<!-- # afs -> soc -->

<!-- # agc -> soc -->

<!-- # climate -> agc -->

<!-- # climate -> soc -->

<!-- # soil -> agc -->

<!-- # soil -> soc -->

<!-- # } -->

>>>>

journal: "Agroforestry systems"
date: ""
output:
  officedown::rdocx_document: 
    base_format: "bookdown::word_document2"
    reference_docx: misc/word-styles-ref-02.docx
bibliography: references.bib
nocite: |
  @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{r setup, include=FALSE}
# Run with R version 4.4.1
# package management is done with renv
packages_needed <- c("maps", "car")
library(ggplot2)
library(ggrepel)
library(ggspatial)
library(ggpubr)
library(ggfortify)
library(BIOMASS)
library(data.table)
library(flextable)
library(terra)
library(rdryad)
library(knitr)
library(readxl)
library(dataverse)

@@

```{r tab.id = "abbrev", results = "show", tab.cap = "List of abbreviations used in the paper."}
data.frame(
  Abbreviation = c("AFS", "AGC", "BGC", "CF", "FG", "HG", "SOC", "SP", "YA"),
  Definition = c(
    "Agroforestry system", "Aboveground carbon", "Belowground carbon",
    "Crop and fallow", "Forest garden",
    "Home garden", "Soil organic carbon", "Silvopastoral",
    "Young agroforest"
  )
) |>
  flextable(cwidth = c(1, 2))
```

@@
